Automated Verification for Real-Time Systems using Implicit Clocks and an Extended Antimirov Algorithm

### 2022-04-14 paper/tv.tex
m

The TRS  is an automated entailment checker to prove language 
inclusions among \timedEffects\ (cf. \defref{Inclusion} and 
\tabref{tab:rewriting_tree_send}). It is triggered i) prior to 
temporal property assertions; ii) prior to function calls for the 
precondition checking; and iii) at the end of verifying a function 
for the post condition checking. Given two effects 
\code{\effect_1,\ \effect_2}, TRS decides if the inclusion \code{\effect_1 \CONTAIN  \effect_2 } is valid.

@@

Now, the inclusions are disjunction-free formulas. 
Next we provide the definitions and the key implementations\footnote{As 
the implementations according to basic regular expressions can be found 
in prior work \cite{DBLP:conf/fsttcs/KeilT14}. 
%Here, we focus on presenting 
%the definitions and how do we deal with dependent values in the effects, 
%as the key novelties of this work. 
%More detail implementations can be found in the extended technical report\citeref{Demo}.
} of Nullable, First and Derivative respectively. Intuitively, 
the Nullable function \code{\delta_{{\pi}}(\es)} returns a boolean 
value indicating whether \code{{\pi} \wedge \es} contains the empty 
trace; the First function \code{fst_{{\pi}}( \es)} computes a set of 
 initial \emph{heads}, denoted as \code{h},  of \code{{\pi} \wedge \es}; and the Derivative 
function \code{D^{\env}_{h}(\pi \wedge \es)} computes a next-state 
effects after eliminating the head  \code{h} from the current 
effects \code{{\pi} \wedge \es}, w.r.t \code{\env}.

@@

\begin{definition}[Nullable]\label{Nullable}
Given any \code{\effect {=} \pi  \wedge  \es}, %we define 
%\code{\delta_{{\pi}}(\es)} to be \\
{
  \code{\delta_{{\pi}}(\es) :bool {=}
  \begin{cases}
        \code{true} & \code{if\  \epsilon \in \llbracket {\pi}  \wedge  \es \rrbracket_{\varphi}}\\
        \code{false} & \code{if\  \epsilon \notin \llbracket {\pi}  \wedge  \es \rrbracket_{\varphi}}
      \end{cases} }
      }, where:  
\begin{comment}
 a boolean value, indicating whether \code{\effect} contains the empty trace, i.e., $\epsilon$. 
We define recursively:  
\end{comment} 
{ 
\begin{gather*}
\code{\delta_{\pi}(\bot ) {=} false} 
\qquad\quad
\code{\delta_{\pi}( \epsilon) {=} true} 
\qquad\quad
\code{\delta_{\pi}(\ev){=} false}   
\qquad\quad
\code{\delta_{\pi}(\pi^\prime? \es){=} \delta_{\pi}(\es)}   
\qquad\quad\
    \code{\delta({\es^\star} ) {=} true}   
\\ 
\code{\delta_{\pi}(\es \cdot \es_2) {=} \delta(\es_1) \wedge \delta(\es_2)}
\qquad\qquad\quad
  \code{\delta_{\pi}(\es_1 \vee \es_2) {=} \delta(\es_1) \vee \delta(\es_2)}   
  \\
\code{\delta_{\pi}(\es_1 || \es_2) {=} \delta(\es_1) \wedge \delta(\es_2)}
\qquad\qquad\quad
 \code{\delta_{\pi}(\es \mysharp  t) {=} SAT ({\pi} \wedge (t {=} 0) )}
\end{gather*}}
\end{definition}

@@

\begin{definition}[Heads]\label{Heads}
If \code{h} is a head of \code{\pi \wedge \es}, then there exists 
\code{\pi^\prime} and \code{\es^\prime}, 
such that \code{\pi \wedge \es} {=} \code{\pi^\prime \wedge h \cdot \es^\prime}. 
   %We adopt the following naming convention for transition labels: 
A head can be  $t$, denoting a pure time passing; 
  $\ev $, denoting a pure event passing; 
  $(\ev, t)$, denoting an event passing which takes time \code{t}. 
\end{definition}

\begin{definition}[First]\label{First}
  Given any \code{\effect {=} \pi \wedge \es},
  \code{fst_{{\pi}}(\es)} return a set of heads, be the set of initial elements   
  derivable from effects \code{\pi \wedge \es}, where: 
    { 
   \begin{gather*} 
  \code{fst_{{\pi}}( \bot ) {=} \{\} } \qquad\ \ 
  \code{fst_{{\pi}}( \epsilon) {=}  \{\} } \qquad\ \ 
  \code{fst_{{\pi}}( \ev) {=} \{ \ev \}  }
  \qquad\ \ 
  \code{fst_{{\pi}}(  \epsilon \mysharp  t) {=}  \{ t \}}
  \qquad\ \ 
  \code{fst_{{\pi}}( {\es^\star} ) {=} fst_{{\pi}}({ \es})}
   \\
   \code{fst_{{\pi}}(\es \mysharp  t) {=}  \{(\ev, t) \ | \ \ev \in fst_{\pi}(\es) \}} 
   \ \ \  \emph{(t\ is\ fresh)}
  \qquad\quad 
  \qquad\quad
  \code{fst_{{\pi}}(  \es_1 {\vee} \es_2) {=} fst_{{\pi}}(  \es_1) \cup fst_{{\pi}}(  \es_2)  }  \\
  \code{fst_{{\pi}}(\epsilon \mysharp  t) {=}  \{t\}} 
   \qquad \qquad
  \code{fst_{{\pi}}( \pi^\prime? \es ) {=} fst_{\pi}(\es)  }
      \qquad\qquad
     \code{fst_{{\pi}}(  \es_1 {||} \es_2) {=} fst_{{\pi}}(  \es_1) \cup fst_{{\pi}}(  \es_2)  }   
     \\
  {
  \code{fst_{{\pi}}(\es_1 {\cdot} \es_2) {=} } 
  \begin{cases}
        \code{\code{fst_{{\pi}}(  \es_1)} \cup \code{fst_{{\pi}}(  \es_2)} } & \code{ if\ \delta(\es_1) {=} true}\\
        \code{\code{fst_{{\pi}}(  \es_1)}} & \code{  if\ \delta(\es_1) {=} false}
      \end{cases} 
      }  
  \end{gather*}
  }
\end{definition}

@@

\begin{definition}[Partial Derivative\footnote{Intuitively, the partial derivative refers to the left quotient of a language equation, for example, for REs, \code{x^{{-}1} \llbracket x\cdot y\rrbracket {=} y};  \code{y^{{-}1} \llbracket x\cdot y\rrbracket {=} \bot }; and  \code{y^{{-}1} \llbracket x + y\rrbracket {=} \epsilon}. Here we come up with a new notion of partial derivative for \timedEffects. 
}]\label{Derivative}
Given any \code{\effect {=} \pi \wedge \es}, and a variable valuation function \code{\env}, 
the partial derivative \code{D^{\env}_{h}(\pi \wedge \es)} 
 computes the  effects for the 
left quotient \code{(h)^{\text{-}1}\llbracket \pi \wedge \es 
\rrbracket_{\env}}. 
%\footnote{As the reader may notice, our First and Derivative function does not handle the timed sequences constructed by \code{||}, as the synchronous parallel is reduced during the normalization phase.}
\begin{gather*}
\code{D^{\env}_{h}(\pi \wedge \bot ) {=}}  \code{False \wedge \bot } 
\qquad\ \ \  
\code{D^{\env}_{h}( \pi \wedge  \epsilon) {=}}  \code{False \wedge \bot } 
\qquad\ \ \ 
\code{D^{\env}_{h}( \pi \wedge {\es^\star}) } {=} \code{ D^{\env}_{h}(\pi \wedge  {\es}) \cdot {\es^\star}}  
\\
{
 \code{\code{D^{\env}_{h}(\pi \wedge \pi^\prime? \es)} {=}}
\begin{cases}
      \code{\pi \wedge \pi^\prime? \es} \quad &\code{if\ \llbracket \pi^\prime  \rrbracket_{\env} {=} false}   \\
      \code{D^{\env}_{h}(\pi \wedge \es)} & \code{if\ \llbracket \pi^\prime  \rrbracket_{\env} {=} true} 
    \end{cases} 
    }
    \\ 
{
 \code{D^{\env}_{h}(\pi \wedge \es_1  \cdot  \es_2) {=} }
\begin{cases}
      \code{D^{\env}_{h}(\pi \wedge \es_1) \cdot  \es_2\ \vee \  D^{\env}_{h}(\pi \wedge \es_2)} \quad &\code{if\ \delta_{\pi}(\es_1) {=} true}   \\
      \code{D^{\env}_{h}(\pi \wedge \es_2) \cdot  \es_2} & \code{if\ \delta_{\pi}(\es_1) {=} false} 
    \end{cases} 
    }
\\
\code{D^{\env}_{h}(\pi \wedge  \es_1 {\vee} \es_2) } {=} \code{ D^{\env}_{h}(\pi \wedge  \es_1) \vee D^{\env}_{h}( \pi \wedge \es_2)} 
\qquad 
\code{D^{\env}_{h}(\pi \wedge \es_1 {||}\es_2)} {=} \code{PD^{\env}_{h}(\pi \wedge \es_1)} || \code{PD^{\env}_{h}(\pi \wedge \es_2)} 
 \\~\\
 \hline
 ~\\
 \code{D^{\env}_{t}(\pi \wedge  \ev) } {=} \code{False \wedge \bot } 
 \qquad\qquad 
 \code{D^{\env}_{t}(\pi \wedge  \epsilon \mysharp t^\prime) } {=}  
 \code{\pi \wedge (t{=}t^\prime) \wedge \epsilon}
 \\
 \code{D^{\env}_{t}(\pi \wedge  \es \mysharp t^\prime) } {=} 
 \code{\pi \wedge (t{+}t^{\prime\prime}{=}t^\prime) \wedge  
 \es \mysharp t^{\prime\prime} \ \ \  {(t^{\prime\prime}\ is\ fresh)} } 
\\~\\
\hline
~\\
\code{D^{\env}_{\ev}(\pi \wedge  \ev^\prime) } {=} 
\code{\pi \wedge (\ev{=}\ev^\prime) \wedge \epsilon} 
\qquad \quad 
\code{D^{\env}_{\ev}(\pi \wedge  \es \mysharp t^\prime) } {=} 
\code{\{(\pi^\prime \wedge \es^\prime \mysharp t^\prime) \ | \ (\pi^\prime \wedge \es^\prime) \in D^{\env}_{\ev}(\pi \wedge  \es)\}} 
\\~\\
\hline
~\\
\code{D^{\env}_{(\ev, t)}(\pi \wedge  \ev^\prime) } {=} 
\code{False \wedge \bot }  
\qquad \ \ \  
\code{D^{\env}_{(\ev, t)}(\pi \wedge  \es \mysharp t^\prime) } {=} 
\code{\pi \wedge (t{+}t^{\prime\prime}{=}t^\prime) \wedge  
(D^{\env}_{\ev}(\pi \wedge  \es)) \mysharp t^{\prime\prime} \ \ \  {(t^{\prime\prime}\ is\ fresh)}} 
\\
\end{gather*}
\end{definition}

\begin{definition}[Parallel Derivative]\label{Derivative_maybe}
  Given any \code{\effect {=} \pi \wedge \es}, and a variable valuation function \code{\env}, 
  the parallel derivative is defined as: {
    \code{PD^{\env}_{h}(\pi \wedge \es) {=}
    \begin{cases}
          \code{\pi \wedge \es} & \code{if\ D^{\env}_{h}(\pi \wedge \es) = (False \wedge \bot )}\\
          \code{D^{\env}_{h}(\pi \wedge \es)} & \code{otherwise}
        \end{cases} }
        }

\end{definition}

>>>>

The TRS  is an automated entailment checker to prove language 
inclusions between \timedEffects, cf., \defref{Inclusion} and 
\tabref{tab:rewriting_tree_send} It is triggered i) prior to 
temporal property assertions; ii) prior to function calls for the 
precondition checking; and iii) at the end of verifying a function 
for the post condition checking. Given two effects 
\code{\effect_1} and \code{\effect_2}, TRS decides if the inclusion \code{\effect_1 \CONTAIN  \effect_2 } is valid.

@@

Now, the inclusions are disjunction-free formulas. 
Next we provide the definitions and the key implementations\footnote{As 
the implementations according to basic regular expressions can be found 
in prior work \cite{DBLP:conf/fsttcs/KeilT14}. 
%Here, we focus on presenting 
%the definitions and how do we deal with dependent values in the effects, 
%as the key novelties of this work. 
%More detail implementations can be found in the extended technical report\citeref{Demo}.
} of Nullable, First and Derivative respectively. Intuitively, 
the Nullable function \code{\delta_{{\pi}}(\es)} returns a boolean 
value indicating whether \code{{\pi} \wedge \es} contains the empty 
trace; the First function \code{fst_{{\pi}}( \es)} computes a set of 
 initial \emph{heads}, denoted as \code{h},  of \code{{\pi} \wedge \es}; and the Derivative 
function \code{D^{\pi}_{h}(\es)} computes a next-state 
effects after eliminating the head  \code{h} from the current 
effects \code{{\pi} \wedge \es}.

@@

\begin{definition}[Nullable \footnote{The proof obligations are 
  discharged using the Z3 SMT prover, while deciding the nullability of effects constructed by symbolic terms, represented by \code{SAT({\pi})}.}]\label{Nullable}
Given any \code{\effect {=} \pi  \wedge  \es}, %we define 
%\code{\delta_{{\pi}}(\es)} to be \\
{
  \code{\delta_{{\pi}}(\es) :bool {=}
  \begin{cases}
        \code{true} & \code{if\  \epsilon \in \llbracket {\pi}  \wedge  \es \rrbracket}\\
        \code{false} & \code{if\  \epsilon \notin \llbracket {\pi}  \wedge  \es \rrbracket}
      \end{cases} }
      }, where:  
\begin{comment}
 a boolean value, indicating whether \code{\effect} contains the empty trace, i.e., $\epsilon$. 
We define recursively:  
\end{comment} 
{ 
\begin{gather*}
\code{\delta_{\pi}(\bot ) {=} false} 
\qquad\quad
\code{\delta_{\pi}( \epsilon) {=} true} 
\qquad\quad
\code{\delta_{\pi}(\ev){=} false}   
\qquad\quad
\code{\delta_{\pi}(\pi^\prime?){=} false}  
\qquad\quad\
    \code{\delta({\es^\star} ) {=} true}   
\\ 
\ \ 
\code{\delta_{\pi}(\es \cdot \es_2) {=} \delta(\es_1) \wedge \delta(\es_2)}
\qquad\qquad\quad
  \code{\delta_{\pi}(\es_1 \vee \es_2) {=} \delta(\es_1) \vee \delta(\es_2)}   
  \\
\code{\delta_{\pi}(\es_1 || \es_2) {=} \delta(\es_1) \wedge \delta(\es_2)}
\qquad\qquad\quad
 \code{\delta_{\pi}(\es \mysharp  t) {=} SAT ({\pi} \wedge (t {=} 0) )}
\end{gather*}}
\end{definition}

@@

\begin{definition}[Heads]\label{Heads}
If \code{h} is a head of \code{\pi \wedge \es}, then there exists 
\code{\pi^\prime} and \code{\es^\prime}, 
such that \code{\pi \wedge \es} {=} \code{\pi^\prime \wedge h \cdot \es^\prime}. 
   %We adopt the following naming convention for transition labels: 
A head can be  $t$, denoting a pure time passing; 
  $\ev $, denoting an instant event passing; or 
  $(\ev, t)$, denoting an event passing which takes time \code{t}. 
\end{definition}

\begin{definition}[First]\label{First}
  Given any \code{\effect {=} \pi \wedge \es},
  \code{fst_{{\pi}}(\es)} return a set of heads, be the set of initial elements   
  derivable from effects \code{\pi \wedge \es}, where: 
    { 
   \begin{gather*} 
  \code{fst_{{\pi}}( \bot ) {=} \{\} } \qquad\quad 
  \code{fst_{{\pi}}( \epsilon) {=}  \{\} } \qquad\quad
  \code{fst_{{\pi}}( \ev) {=} \{ \ev \}  }
  \qquad\quad
  \code{fst_{{\pi}}(  \epsilon \mysharp  t) {=}  \{ t \}}
  \qquad\quad
  \code{fst_{{\pi}}( {\es^\star} ) {=} fst_{{\pi}}({ \es})}
   \\
   \code{fst_{{\pi}}(\es \mysharp  t) {=}  \{(\ev, t^\prime) \ | \ \ev \in fst_{\pi}(\es) \}} 
   \ \ \  \emph{($t^\prime$\ is\ fresh)}
  \quad  
  \qquad
  \code{fst_{{\pi}}(  \es_1 {\vee} \es_2) {=} fst_{{\pi}}(  \es_1) \cup fst_{{\pi}}(  \es_2)  }  
  \\
  \code{fst_{{\pi}}( \pi^\prime? ) {=} \{\tau(\pi^\prime); \tau(\neg \pi^\prime)\}  }
      \qquad\qquad\qquad 
     \code{fst_{{\pi}}(  \es_1 {||} \es_2) {=} fst_{{\pi}}(  \es_1) \cup fst_{{\pi}}(  \es_2)  }   
     \\
  {
  \code{fst_{{\pi}}(\es_1 \cdot \es_2) {=} } 
  \begin{cases}
        \code{\code{fst_{{\pi}}(  \es_1)} \cup \code{fst_{{\pi}}(  \es_2)} } \quad   & \code{ if\ \  \delta(\es_1) {=} true}\\
        \code{\code{fst_{{\pi}}(  \es_1)}} & \code{  if\ \  \delta(\es_1) {=} false}
      \end{cases} 
      }  
  \end{gather*}
  }
\end{definition}

@@

\begin{definition}[Partial Derivative]\label{Derivative}
  Given any \code{\effect {=}\pi \wedge  \es}, 
  %and a variable valuation function \code{\env}, 
  the partial derivative \code{D^{\pi}_{h}( \es)} 
   computes the  effects for the 
  left quotient \code{(h)^{\text{-}1}\llbracket \pi \wedge \es 
  \rrbracket}\footnote{Intuitively, the partial derivative refers to the left quotient of a language equation, for example, for REs, \code{x^{{-}1} \llbracket x\cdot y\rrbracket {=} y};  \code{y^{{-}1} \llbracket x\cdot y\rrbracket {=} \bot }; and  \code{y^{{-}1} \llbracket x + y\rrbracket {=} \epsilon}. Here we come up with a new notion of partial derivative for \timedEffects. 
  }. 
  %\footnote{As the reader may notice, our First and Derivative function does not handle the timed sequences constructed by \code{||}, as the synchronous parallel is reduced during the normalization phase.}
  \begin{gather*}
  \code{D^{\pi}_{h}( \bot ) {=}}  \code{False \wedge \bot } 
  \qquad\qquad 
  \code{D^{\pi}_{h}(   \epsilon) {=}}  \code{False \wedge \bot } 
  \qquad\qquad
  \code{D^{\pi}_{h}(  \ev^\prime) } {=} 
  \code{  (\pi \wedge h{=}\ev^\prime) \wedge \epsilon} 
  \\
  {
   \code{\code{D^{\pi}_{h}( \pi^\prime?)} {=}}
  \begin{cases}
        \code{\pi \wedge \pi^\prime? } &\code{if\ h {\not\Rightarrow}  \pi^\prime}   \\
        \code{\pi \wedge \epsilon} & \code{if\ h {\Rightarrow} \pi^\prime} 
      \end{cases} 
      }
      \qquad \quad \ \ \ 
  {
   \code{D^{\pi}_{h}( \es_1  \cdot  \es_2) {=} }
  \begin{cases}
        \code{D^{\pi}_{h}( \es_1) \cdot  \es_2\ \vee \  D^{\pi}_{h}( \es_2)}  &\code{if\ \delta_{\pi}(\es_1) {=} true}   \\
        \code{D^{\pi}_{h}( \es_2) \cdot  \es_2} & \code{if\ \delta_{\pi}(\es_1) {=} false} 
      \end{cases} 
      }
  \\
  \code{D^{\pi}_{h}(  {\es^\star}) } {=} \code{ D^{\pi}_{h}(  {\es}) \cdot {\es^\star}}  
  \qquad\quad \ \ \ 
  \code{D^{\pi}_{h}(  \es_1 {\vee} \es_2) } {=} \code{ D^{\pi}_{h}(  \es_1) \vee D^{\pi}_{h}(  \es_2)} 
  \qquad \quad\ \ \ 
  \code{D^{\pi}_{h}( \es_1 {||}\es_2)} {=} \code{\Bar{\Bar{D}}^{\pi}_{h}( \es_1)} || \code{\Bar{\Bar{D}}^{\pi}_{h}( \es_2)} 
   \\
   \code{D^{\pi}_{\ev}(  \es \mysharp t) } {=} {\bigvee}
   \code{ \{ (\pi^\prime \wedge \es^\prime  \mysharp t) \ | \ (\pi^\prime, \es^\prime) \in D^{\pi}_{\ev}(\es) \} } 
\qquad 
   \code{D^{\pi}_{t}(  \es \mysharp t^\prime) } {=} 
   \code{ (\pi \wedge t{+}t^{\prime\prime}{=}t^\prime) \wedge  
   \es \mysharp t^{\prime\prime} \ \ \  {(t^{\prime\prime}\ is\ fresh)} } 
  \end{gather*}
  \end{definition}

Notice that the derivatives of a parallel composition makes use of the \emph{Parallel Derivative} \code{\Bar{\Bar{D}}^{\pi}_{h}(\es)}, defined as follows: 
  \begin{definition}[Parallel Derivative]\label{Derivative_maybe}
    Given any \code{\effect {=} \pi \wedge \es}, and a variable valuation function \code{\env}, 
    the parallel derivative is defined as: {
      \code{\Bar{\Bar{D}}^{\pi}_{h}( \es) {=}
      \begin{cases}
            \code{ \es} & \code{if\ D^{\pi}_{h}(\pi \wedge \es) = (False \wedge \bot )}\\
            \code{D^{\pi}_{h}( \es)} & \code{otherwise}
          \end{cases} }
    }.
  \end{definition}
